---
title: "Portfolio 2: Thesis Part 2 (State-Level Analyses)"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```


This is part 2 of the analyses for my master's thesis project, which was titled "Effects of a Short-Term Volitional Intervention Promoting Moral and Non-Moral Personality Change." Here is the abstract of my thesis for some context:

> People can intentionally change specific personality traits. However, it is unclear whether such interventions are effective at changing moral character traits. The present study (total N = 179), which utilized two active intervention groups for mutual comparisons, examined whether a 5-week goal-setting intervention could lead to trait change for a moral trait, compassion (n = 87), or the Big 5 trait of conscientiousness (n = 92). Participants self-selected into one condition after getting feedback on where they stood on these traits in relation to the population norm and completing a brief reflection task. They set implementation intentions and reflected on their progress every 3 days for the duration of the intervention. State assessments were collected every 3 days in addition to self- and observer-report trait assessments at pre- and post-test, with follow-up self-report assessments 2 weeks post-program. Results from analyses of self-report trait and state assessments indicated that participants in both intervention groups were able to successfully change their personalities in the short term, providing preliminary evidence that implementation intention-based volitional personality change interventions can be useful tools for both moral and non-moral trait change. 

### Loading Packages and Data

```{r load-packages-data}

library(foreign)

VPCcombined = read.spss("VPCcombinedclean_3.26.23 with covariates_copy.sav",
                      to.data.frame = TRUE,
                      reencode = TRUE,
                      use.value.labels = FALSE)

library(nlme)
library(lme4)
library(lmerTest)
library(psych)
library(multilevel)
library(tidyverse)

VPCstate <- filter(VPCcombined, datalevel == 2)

VPCstate$id <- as.factor(VPCstate$id)
VPCstate$intervention <- as.factor(VPCstate$intervention)

ConsState <- filter(VPCstate, intervention == 1)
CompState <- filter(VPCstate, intervention == 0)


```


### Hypothesis 5 

Hypothesis 5: I hypothesized that those in the **conscientiousness** group will experience significant ***increases levels in state conscientiousness*** over the duration of the study. 

Starting with an intercept only model:

```{r state-conscienctiousness-null, message=FALSE, warning=FALSE}
StateConsNull <- lmer(stateconscientiousness ~  1 +
                (1 | id), 
              data = ConsState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateConsNull)
confint(StateConsNull)
rand(StateConsNull)
AIC(StateConsNull)
BIC(StateConsNull)
```

Then linear:

```{r state-conscientiousness-linear, message=FALSE, warning=FALSE}
StateConsLin <- lmer(stateconscientiousness ~  StateLinearTime +
                (1 + StateLinearTime | id), 
              data = ConsState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateConsLin)
confint(StateConsLin)
rand(StateConsLin)
AIC(StateConsLin)
BIC(StateConsLin)
```

Then log:

```{r state-conscientiousness-log, message=FALSE, warning=FALSE}
StateConsLog <- lmer(stateconscientiousness ~  StateLogTime +
                (1 + StateLogTime | id), 
              data = ConsState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateConsLog)
confint(StateConsLog)
rand(StateConsLog)
AIC(StateConsLog)
BIC(StateConsLog)
```

Then with pretest conscientiousness covariate:

```{r state-conscienctiousness-log-with-covariate, message=FALSE, warning=FALSE}
StateConsLogCo <- lmer(stateconscientiousness ~ pretestcons + StateLogTime +
                (1 + StateLogTime | id), 
              data = ConsState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateConsLogCo)
confint(StateConsLogCo)
rand(StateConsLogCo)
AIC(StateConsLogCo)
BIC(StateConsLogCo)
```

Then examining patterns in state compassion for those working on their conscientiousness:

```{r state-conscienctiousness-log-nontarget-state, message=FALSE, warning=FALSE}
StateConsNonTarget <- lmer(statecompassion ~  StateLogTime +
                (1 + StateLogTime | id), 
              data = ConsState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateConsNonTarget)
confint(StateConsNonTarget)
rand(StateConsNonTarget)
AIC(StateConsNonTarget)
BIC(StateConsNonTarget)
```

#### Conclusion:

>Hypothesis 5 was supported: participants in the conscientiousness group experienced significant increases in state conscientiousness over the duration of the study (b = 0.08, 95% CI [0.01, 0.14], SE = 0.03, p < .05). There was a medium-large effect for state conscientiousness for participants working on their conscientiousness (d = 0.61). However, those in the conscientiousness group also increased in state compassion (b = 0.07, 95% CI [0.03, 0.12], SE = 0.02, p < .01) with a medium-sized effect (d = 0.48). 


### Hypothesis 6 

Hypothesis 6: I hypothesized that those in the **compassion** group will experience significant ***increases levels in state compassion*** over the duration of the study. 


Starting with an intercept only model:

```{r state-compassion-null, message=FALSE, warning=FALSE}
StateCompNull <- lmer(statecompassion ~ 1 +
                (1 | id), 
              data = CompState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateCompNull)
confint(StateCompNull)
rand(StateCompNull)
AIC(StateCompNull)
BIC(StateCompNull)
```

Then linear:

```{r state-compassion-linear, message=FALSE, warning=FALSE}
StateCompLin <- lmer(statecompassion ~  StateLinearTime +
                (1 + StateLinearTime | id), 
              data = CompState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateCompLin)
confint(StateCompLin)
rand(StateCompLin)
AIC(StateCompLin)
BIC(StateCompLin)
```

Then log: 

```{r state-compassion-log, message=FALSE, warning=FALSE}
StateCompLog <- lmer(statecompassion ~  StateLogTime +
                (1 + StateLogTime | id), 
              data = CompState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateCompLog)
confint(StateCompLog)
rand(StateCompLog)
AIC(StateCompLog)
BIC(StateCompLog)
```

(Log had the best fit again.)


With the same covariate as before:

```{r state-compassion-log-with-covariate, message=FALSE, warning=FALSE}
StateCompLogCo <- lmer(statecompassion ~ pretestcons + StateLogTime +
                (1 + StateLogTime | id), 
              data = CompState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateCompLogCo)
confint(StateCompLogCo)
rand(StateCompLogCo)
AIC(StateCompLogCo)
BIC(StateCompLogCo)
```

Examining patterns in non-target trait at the state level:

```{r state-compassion-log-nontarget-state, message=FALSE, warning=FALSE}
StateCompNonTarget <- lmer(stateconscientiousness ~  StateLogTime +
                (1 + StateLogTime | id), 
              data = CompState,
              REML = FALSE,
              na.action = "na.omit")

summary(StateCompNonTarget)
confint(StateCompNonTarget)
rand(StateCompNonTarget)
AIC(StateCompNonTarget)
BIC(StateCompNonTarget)
```

#### Conclusion:

>Hypothesis 6 was also supported: those in the compassion group experienced significant increases in state compassion over the duration of the study (b = 0.16, 95% CI [0.11, 0.20], SE = 0.02, p < .001). There was a large effect for state compassion for compassion group participants (d = 0.70). Those in the compassion group did not experience a significant increase in levels of state conscientiousness (b = 0.02, 95% CI [-0.02, 0.08], SE = 0.03, p = .37). 



### Hypothesis 7

I examined the intervention groups combined to detect Time x Group interaction effects at the state level.

Hypothesis 7: I hypothesized that **people who chose the conscientiousness intervention** will show ***greater increases in state conscientiousness compared with the compassion group.***

Testing the group by time (slope) interaction:

Null first:

```{r state-conscientiousness-group-by-slope-null, message=FALSE, warning=FALSE}
ConsInteractionNull <- lmer(stateconscientiousness ~  1 +
                (1 | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionNull)
confint(ConsInteractionNull)
rand(ConsInteractionNull)
AIC(ConsInteractionNull)
BIC(ConsInteractionNull)
```

Then linear:

```{r state-conscientiousness-group-by-slope-linear, message=FALSE, warning=FALSE}
ConsInteractionLin <- lmer(stateconscientiousness ~  StateLinearTime*intervention +
                (1 + StateLinearTime | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionLin)
confint(ConsInteractionLin)
rand(ConsInteractionLin)
AIC(ConsInteractionLin)
BIC(ConsInteractionLin)
```


Then log:

```{r state-conscientiousness-group-by-slope-log, message=FALSE, warning=FALSE}
ConsInteractionLog <- lmer(stateconscientiousness ~  StateLogTime*intervention +
                (1 + StateLogTime | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionLog)
confint(ConsInteractionLog)
rand(ConsInteractionLog)
AIC(ConsInteractionLog)
BIC(ConsInteractionLog)
```

Then adding pretest conscientiousness covariate:

```{r state-conscientiousness-group-by-slope-log-with-covariate, message=FALSE, warning=FALSE}
ConsInteractionCo <- lmer(stateconscientiousness ~ pretestcons + StateLogTime*intervention +
                (1 + StateLogTime | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionCo)
confint(ConsInteractionCo)
rand(ConsInteractionCo)
AIC(ConsInteractionCo)
BIC(ConsInteractionCo)
```

#### Conclusion:

>Hypothesis 7 was not fully supported: although participants in the conscientiousness group experienced greater increases in state conscientiousness than the compassion group’s increases in state conscientiousness (b = 0.06, 95% CI [-0.03, 0.14], SE = 0.04), this effect was not significant (p = .09). Although nonsignificant, the conscientiousness group experienced a 0.08 unit change in state conscientiousness in the expected direction, while the compassion group did not experience a significant increase in state conscientiousness (b = 0.02, p = .46).he logarithmic slope by group interactions still held when controlling for pretest levels of trait conscientiousness.


### Hypothesis 8

Hypothesis 8: I hypothesized that **people who chose the compassion intervention** will show ***greater increases in state compassion compared with the conscientiousness group.***

Testing the group by time (slope) interaction:

```{r state-compassion-group-by-slope-null, message=FALSE, warning=FALSE}
CompInteractionNull <- lmer(statecompassion ~  1 +
                (1 | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionNull)
confint(CompInteractionNull)
rand(CompInteractionNull)
AIC(CompInteractionNull)
BIC(CompInteractionNull)
```


Testing a linear model:

```{r state-compassion-group-by-slope-linear, warnings=FALSE, message=FALSE, warning=FALSE}
CompInteractionLin <- lmer(statecompassion ~  StateLinearTime*intervention +
                (1 + StateLinearTime | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionLin)
confint(CompInteractionLin)
rand(CompInteractionLin)
AIC(CompInteractionLin)
BIC(CompInteractionLin)
```

Testing a logarithmic model:

```{r state-compassion-group-by-slope-log, message=FALSE, warning=FALSE}
CompInteractionLog <- lmer(statecompassion ~  StateLogTime*intervention +
                (1 + StateLogTime | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionLog)
confint(CompInteractionLog)
rand(CompInteractionLog)
AIC(CompInteractionLog)
BIC(CompInteractionLog)
```

Adding the conscientiousness covariate:

```{r state-compassion-group-by-slope-log-with-covariate, message=FALSE, warning=FALSE}
CompInteractionCo <- lmer(statecompassion ~ pretestcons + StateLogTime*intervention +
                (1 + StateLogTime | id), 
              data = VPCstate,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionCo)
confint(CompInteractionCo)
rand(CompInteractionCo)
AIC(CompInteractionCo)
BIC(CompInteractionCo)
```

#### Conclusion:

>Hypothesis 8 was supported: the logarithmic slope by group effect for state compassion (b = -0.09, 95% CI [-0.15, -0.02], SE = 0.03, p < .01) was significant. The compassion group experienced a significant 0.15 unit increase in state-level compassion (p < .001), while the conscientiousness group experienced a smaller 0.06 (0.15 – 0.09) unit increase in state compassion (p < .01). The logarithmic slope by group interactions still held when controlling for pretest levels of trait conscientiousness.



