---
title: "Portfolio 1: Thesis Part 1 (Trait-Level Analyses)"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```


This is part 1 of the analyses for my master's thesis project, which was titled "Effects of a Short-Term Volitional Intervention Promoting Moral and Non-Moral Personality Change." Here is the abstract of my thesis for some context:

> People can intentionally change specific personality traits. However, it is unclear whether such interventions are effective at changing moral character traits. The present study (total N = 179), which utilized two active intervention groups for mutual comparisons, examined whether a 5-week goal-setting intervention could lead to trait change for a moral trait, compassion (n = 87), or the Big 5 trait of conscientiousness (n = 92). Participants self-selected into one condition after getting feedback on where they stood on these traits in relation to the population norm and completing a brief reflection task. They set implementation intentions and reflected on their progress every 3 days for the duration of the intervention. State assessments were collected every 3 days in addition to self- and observer-report trait assessments at pre- and post-test, with follow-up self-report assessments 2 weeks post-program. Results from analyses of self-report trait and state assessments indicated that participants in both intervention groups were able to successfully change their personalities in the short term, providing preliminary evidence that implementation intention-based volitional personality change interventions can be useful tools for both moral and non-moral trait change. 

For all of my thesis analyses (parts/portfolios 1-4) my code was modeled after the code of [Stieger et al. (2020)](https://journals.sagepub.com/doi/pdf/10.1002/per.2267?casa_token=Jq_sdyLlvDkAAAAA:voEE-qLnu6Vf51AigHKGN4H725UJPD-W00Cvr4GNjAcZdMBboRB47bF-RX4crih5MAiOMvOHmTSy), which can be found [here](https://osf.io/an5dj/). Before running my own analyses, I first re-ran all of their code with their data and matched it with the results reported in their paper (everything worked out as expected).

### Loading Packages and Data

```{r load-packages-data}

library(foreign)

VPCcombined = read.spss("VPCcombinedclean_3.26.23 with covariates_copy.sav",
                      to.data.frame = TRUE,
                      reencode = TRUE,
                      use.value.labels = FALSE)

library(nlme)
library(lme4)
library(lmerTest)
library(psych)
library(multilevel)
library(tidyverse)

VPCtrait <- filter(VPCcombined, datalevel == 1)

VPCtrait$id <- as.factor(VPCtrait$id)
VPCtrait$intervention <- as.factor(VPCtrait$intervention)

ConsTrait <- filter(VPCtrait, intervention == 1)
CompTrait <- filter(VPCtrait, intervention == 0)



```



----------------


*From my thesis prospectus:*

"For the trait data (from wave 1, pretest, wave 11, posttest, and wave 12, follow-up), modeling the analytic approach of Stieger et al. (2020; 2021), I will conduct a series of longitudinal multilevel models (with Level 1 as Time, from 0-2, and Level 2 as Person) to detect intervention effects. I will first conduct analyses for the conscientiousness and compassion groups separately (to test Hypotheses 1 and 2, respectively) and then I will examine the groups combined and test for Time x Group interactions (to test Hypotheses 3 and 4)."


"For all analyses, I will start with an intercept-only model to see the pattern of change over time. Then, based on what I see, I will fit different models of change to the data. With the best-fitting model, I will examine the outcome variables of self-reported compassion and conscientiousness levels through follow-up and report fit indices and effect sizes."


### Hypothesis 1 

Hypothesis 1: I hypothesized that those in the **conscientiousness** group will have ***higher levels of trait conscientiousness*** at posttest compared to pretest and that these increases would be retained at follow-up. 

Starting with an intercept only model:

```{r trait-conscienctiousness-null, message=FALSE, warning=FALSE}
TraitConsNull <- lmer(traitconscientiousness ~  1 +
                (1 | id), 
              data = ConsTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitConsNull)
confint(TraitConsNull)
rand(TraitConsNull)
AIC(TraitConsNull)
BIC(TraitConsNull)
```

Then a linear model:

```{r trait-conscientiousness-linear, message=FALSE, warning=FALSE}
TraitConsLin <- lmer(traitconscientiousness ~  TraitLinearTime +
                (1 + TraitLinearTime | id), 
              data = ConsTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitConsLin)
confint(TraitConsLin)
rand(TraitConsLin)
AIC(TraitConsLin)
BIC(TraitConsLin)
```

And finally a log model:

```{r trait-conscientiousness-log, message=FALSE, warning=FALSE}
TraitConsLog <- lmer(traitconscientiousness ~  TraitLogTime +
                (1 + TraitLogTime | id), 
              data = ConsTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitConsLog)
confint(TraitConsLog)
rand(TraitConsLog)
AIC(TraitConsLog)
BIC(TraitConsLog)
```

Of the three models, the logarithmic model has the lowest AIC and BIC values and therefore has the best fit. 


I also took the best-fitting model and added the covariate of pre-test level of trait conscientiousness to see if the effects still held:

```{r trait-conscienctiousness-log-with-covariate, message=FALSE, warning=FALSE}
# TraitConsLogCo <- lmer(traitconscientiousness ~ pretestcons + TraitLogTime +
#                 (1 + TraitLogTime | id), 
#               data = ConsTrait,
#               REML = FALSE,
#               na.action = "na.omit")
# 
# summary(TraitConsLogCo)
# confint(TraitConsLogCo)
# rand(TraitConsLogCo)
# AIC(TraitConsLogCo)
# BIC(TraitConsLogCo)
```

And then I also checked to see if there were any significant change patterns in trait-level compassion for those working on their conscientiousness:

```{r trait-conscienctiousness-log-nontarget-trait, message=FALSE, warning=FALSE}
TraitConsNonTarget <- lmer(traitcompassion ~  TraitLogTime +
                (1 + TraitLogTime | id), 
              data = ConsTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitConsNonTarget)
confint(TraitConsNonTarget)
rand(TraitConsNonTarget)
AIC(TraitConsNonTarget)
BIC(TraitConsNonTarget)
```

#### Conclusion:

Hypothesis 1 was met: those in the **conscientiousness** group had ***significantly higher levels of trait conscientiousness*** at posttest compared to pretest and these increases were generally retained at follow-up. 

Here is the corresponding section from my thesis results section:

>Hypothesis 1 was supported: those in the conscientiousness group had significantly higher levels of trait conscientiousness at posttest compared to pretest and these increases were generally retained at follow-up (b = 0.13, 95% CI [0.07, 0.18], SE = 0.03, p < .001). There was a medium-large effect for trait conscientiousness for participants working on their conscientiousness (d = 0.52). Those in the conscientiousness group did not increase in trait compassion (b = 0.00, 95% CI [-0.03, 0.04], SE = 0.02, p = 0.89). 






### Hypothesis 2 

Hypothesis 2: I hypothesized that those in the **compassion** group will have ***higher levels of trait compassion*** at posttest compared to pretest and that these increases would be retained at follow-up. 

Starting with an intercept only model:

```{r trait-compassion-null, message=FALSE, warning=FALSE}
TraitCompNull <- lmer(traitcompassion ~ 1 +
                (1 | id), 
              data = CompTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitCompNull)
confint(TraitCompNull)
rand(TraitCompNull)
AIC(TraitCompNull)
BIC(TraitCompNull)
```

Then trying linear:

```{r trait-compassion-linear, message=FALSE, warning=FALSE}
TraitCompLin <- lmer(traitcompassion ~  TraitLinearTime +
                (1 + TraitLinearTime | id), 
              data = CompTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitCompLin)
confint(TraitCompLin)
rand(TraitCompLin)
AIC(TraitCompLin)
BIC(TraitCompLin)
```


Then trying log:

```{r trait-compassion-log, message=FALSE, warning=FALSE}
TraitCompLog <- lmer(traitcompassion ~  TraitLogTime +
                (1 + TraitLogTime | id), 
              data = CompTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitCompLog)
confint(TraitCompLog)
rand(TraitCompLog)
AIC(TraitCompLog)
BIC(TraitCompLog)
```

Again, the logarithmic model has the best fit.

Adding the pretest conscientiousness covariate:

```{r trait-compassion-log-with-covariate, message=FALSE, warning=FALSE}
TraitCompLogCo <- lmer(traitcompassion ~ pretestcons + TraitLogTime +
                (1 + TraitLogTime | id), 
              data = CompTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitCompLogCo)
confint(TraitCompLogCo)
rand(TraitCompLogCo)
AIC(TraitCompLogCo)
BIC(TraitCompLogCo)
```

Examining changes in conscientiousness for those working on their compassion:

```{r trait-compassion-log-nontarget-trait, message=FALSE, warning=FALSE}
TraitCompNonTarget <- lmer(traitconscientiousness ~  TraitLogTime +
                (1 + TraitLogTime | id), 
              data = CompTrait,
              REML = FALSE,
              na.action = "na.omit")

summary(TraitCompNonTarget)
confint(TraitCompNonTarget)
rand(TraitCompNonTarget)
AIC(TraitCompNonTarget)
BIC(TraitCompNonTarget)
```


#### Conclusion:

>Hypothesis 2 was also supported: those in the compassion group had significantly higher levels of trait compassion at posttest compared to pretest and these increases were mostly retained at follow-up (b = 0.05, 95% CI [0.01, 0.09], SE = 0.02, p < .05). There was a medium effect for trait compassion for participants working on their compassion (d = 0.28). Those in the compassion group also did not increase in trait conscientiousness (b = 0.01, 95% CI [-0.03, 0.04], SE = 0.02, p = 0.61). 

### Hypothesis 3

In addition to examining the intervention groups separately, I also examined the groups combined to test for Time x Group interactions (i.e., using each group as a control group for the other).

Hypothesis 3: I hypothesized that **people who chose the conscientiousness intervention** will show ***greater increases in trait conscientiousness compared with the compassion group.***

Testing the group by time (slope) interaction:

Null model first:

```{r trait-conscientiousness-group-by-slope-null, message=FALSE, warning=FALSE}
ConsInteractionNull <- lmer(traitconscientiousness ~  1 +
                (1 | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionNull)
confint(ConsInteractionNull)
rand(ConsInteractionNull)
AIC(ConsInteractionNull)
BIC(ConsInteractionNull)
```


Then linear:
```{r trait-conscientiousness-group-by-slope-linear, message=FALSE, warning=FALSE}
ConsInteractionLin <- lmer(traitconscientiousness ~  TraitLinearTime*intervention +
                (1 + TraitLinearTime | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionLin)
confint(ConsInteractionLin)
rand(ConsInteractionLin)
AIC(ConsInteractionLin)
BIC(ConsInteractionLin)
```


Then log: 

```{r trait-conscientiousness-group-by-slope-log, message=FALSE, warning=FALSE}
ConsInteractionLog <- lmer(traitconscientiousness ~  TraitLogTime*intervention +
                (1 + TraitLogTime | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionLog)
confint(ConsInteractionLog)
rand(ConsInteractionLog)
AIC(ConsInteractionLog)
BIC(ConsInteractionLog)
```
Again, the logarithmic model has the best fit.


Adding the pretest level of conscientiousness covariate:

```{r trait-conscientiousness-group-by-slope-log-with-covariate, message=FALSE, warning=FALSE}
ConsInteractionCo <- lmer(traitconscientiousness ~ pretestcons + TraitLogTime*intervention +
                (1 + TraitLogTime | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsInteractionCo)
confint(ConsInteractionCo)
rand(ConsInteractionCo)
AIC(ConsInteractionCo)
BIC(ConsInteractionCo)
```


#### Conclusion:

>Supporting Hypothesis 3, the logarithmic slope by group effect for conscientiousness (b = 0.14, 95% CI [0.07, 0.20], SE = 0.03, p < .001) was significant. The compassion group did not experience a significant increase in trait conscientiousness (p = .90), while the conscientiousness group experienced a 0.14 unit increase in trait conscientiousness (p < .001). The logarithmic slope by group interactions still held when controlling for pretest levels of trait conscientiousness.




### Hypothesis 4

Hypothesis 4: I hypothesized that **people who chose the compassion intervention** will show ***greater increases in trait compassion compared with the conscientiousness group.***

Testing the group by time (slope) interaction:

Testing the null model first—

```{r trait-compassion-group-by-slope-null, message=FALSE, warning=FALSE}
CompInteractionNull <- lmer(traitcompassion ~  1 +
                (1 | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionNull)
confint(CompInteractionNull)
rand(CompInteractionNull)
AIC(CompInteractionNull)
BIC(CompInteractionNull)
```

Testing a linear model:

```{r trait-compassion-group-by-slope-linear, message=FALSE, warning=FALSE}
CompInteractionLin <- lmer(traitcompassion ~  TraitLinearTime*intervention +
                (1 + TraitLinearTime | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionLin)
confint(CompInteractionLin)
rand(CompInteractionLin)
AIC(CompInteractionLin)
BIC(CompInteractionLin)
```

Testing a logarithmic model:

```{r trait-compassion-group-by-slope-log, message=FALSE, warning=FALSE}
CompInteractionLog <- lmer(traitcompassion ~  TraitLogTime*intervention +
                (1 + TraitLogTime | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionLog)
confint(CompInteractionLog)
rand(CompInteractionLog)
AIC(CompInteractionLog)
BIC(CompInteractionLog)
```

Again, the logarithmic model has the best fit. 

With pretest conscientiousness covariate added: 

```{r trait-compassion-group-by-slope-log-with-covariate, message=FALSE, warning=FALSE}
CompInteractionCo <- lmer(traitcompassion ~ pretestcons + TraitLogTime*intervention +
                (1 + TraitLogTime | id), 
              data = VPCtrait,
              REML = FALSE,
              na.action = "na.omit")

summary(CompInteractionCo)
confint(CompInteractionCo)
rand(CompInteractionCo)
AIC(CompInteractionCo)
BIC(CompInteractionCo)
```


#### Conclusion:

>Supporting Hypothesis 4, the logarithmic slope by group effect for compassion (b = -0.05, 95% CI [-0.10, -0.00], SE = 0.02, p < .05) was also significant. The compassion group experienced a small but significant 0.05 unit increase in trait compassion (p < .01), while the conscientiousness group did not experience an increase (0.05 – 0.05) in trait compassion (p < .01). The logarithmic slope by group interactions still held when controlling for pretest levels of trait conscientiousness.

