---
title: 'Portfolio 11: Thesis Part 7 (More Figures)'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is part 7 of the analyses for my master's thesis project, which was titled "Effects of a Short-Term Volitional Intervention Promoting Moral and Non-Moral Personality Change." Here is the abstract of my thesis for some context:

> People can intentionally change specific personality traits. However, it is unclear whether such interventions are effective at changing moral character traits. The present study (total N = 179), which utilized two active intervention groups for mutual comparisons, examined whether a 5-week goal-setting intervention could lead to trait change for a moral trait, compassion (n = 87), or the Big 5 trait of conscientiousness (n = 92). Participants self-selected into one condition after getting feedback on where they stood on these traits in relation to the population norm and completing a brief reflection task. They set implementation intentions and reflected on their progress every 3 days for the duration of the intervention. State assessments were collected every 3 days in addition to self- and observer-report trait assessments at pre- and post-test, with follow-up self-report assessments 2 weeks post-program. Results from analyses of self-report trait and state assessments indicated that participants in both intervention groups were able to successfully change their personalities in the short term, providing preliminary evidence that implementation intention-based volitional personality change interventions can be useful tools for both moral and non-moral trait change.


In portfolios 5 and 6, I went through the process of constructing the main figure to represent my thesis project results in my thesis itself and for various presentations. In this portfolio, I will be documenting my process of attempting to create a more complex figure that conveys the multilevel nature of my data. 


### Loading Packages and Data

```{r load-packages-data}

library(foreign)

VPCcombined = read.spss("VPCcombinedclean_3.26.23 with covariates_copy.sav",
                      to.data.frame = TRUE,
                      reencode = TRUE,
                      use.value.labels = FALSE)

library(nlme)
library(lme4)
library(lmerTest)
library(psych)
library(multilevel)
library(tidyverse)

VPCtrait <- filter(VPCcombined, datalevel == 1)

VPCtrait$id <- as.factor(VPCtrait$id)
VPCtrait$intervention <- as.factor(VPCtrait$intervention)

ConsTrait <- filter(VPCtrait, intervention == 1)
CompTrait <- filter(VPCtrait, intervention == 0)

VPCstate <- filter(VPCcombined, datalevel == 2)

VPCstate$id <- as.factor(VPCstate$id)
VPCstate$intervention <- as.factor(VPCstate$intervention)

ConsState <- filter(VPCstate, intervention == 1)
CompState <- filter(VPCstate, intervention == 0)




```

### Changes in Trait-Level Conscientiousness

Here I'm realizing that you don't have to plot with different datasets to be able to change your legend text!! What a surprise :)

```{r plot-1}

ggplot(VPCtrait, aes(x = TraitLinearTime, y = traitconscientiousness, group = id, color = intervention)) +
  geom_line() +
  labs(title = "Changes in trait-level conscientiousness", subtitle = "by intervention group", x = "Week", y = "Trait Conscientiousness (1-5)", color = "Intervention group") +
   theme(plot.title = element_text(color = "#2E415A", size = 18, face = "bold"),
          plot.subtitle = element_text(color = "#2E415A", size = 16),
          strip.text.x = element_text(color = "#2E415A", size = 16, face = "bold.italic"),
      legend.title = element_text(color = "#2E415A", size = 12, face = "bold"),
      legend.text = element_text(color = "#2E415A", size = 10),
      axis.title = element_text(color = "#2E415A", size = 14),
      axis.text = element_text(color = "#2E415A", size = 12))+
    scale_color_manual(values = c("#E0A151","#B04326"),
                       labels = c("Working on compassion", "Working on conscientiousness"))



```

### Changes in Trait-Level Compassion

Here I'm realizing that you don't have to plot with different datasets to be able to change your legend text!! What a surprise :)

```{r plot-2}

ggplot(VPCtrait, aes(x = TraitLinearTime, y = traitcompassion, group = id, color = intervention)) +
  geom_line() +
  labs(title = "Changes in trait-level compassion", subtitle = "by intervention group", x = "Week", y = "Trait Compassion (1-5)", color = "Intervention group") +
   theme(plot.title = element_text(color = "#2E415A", size = 18, face = "bold"),
          plot.subtitle = element_text(color = "#2E415A", size = 16),
          strip.text.x = element_text(color = "#2E415A", size = 16, face = "bold.italic"),
      legend.title = element_text(color = "#2E415A", size = 12, face = "bold"),
      legend.text = element_text(color = "#2E415A", size = 10),
      axis.title = element_text(color = "#2E415A", size = 14),
      axis.text = element_text(color = "#2E415A", size = 12))+
    scale_color_manual(values = c("#E0A151","#B04326"),
                       labels = c("Working on compassion", "Working on conscientiousness"))



```

Also going to try this plot without the conscientiousness folks to more easily see the patterns, hopefully:

```{r plot-3}

ggplot(CompTrait, aes(x = TraitLinearTime, y = traitcompassion, group = id, color = intervention)) +
  geom_line() +
  labs(title = "Changes in trait-level compassion", x = "Week", y = "Trait Compassion (1-5)", color = "Intervention group") +
   theme(plot.title = element_text(color = "#2E415A", size = 18, face = "bold"),
          plot.subtitle = element_text(color = "#2E415A", size = 16),
          strip.text.x = element_text(color = "#2E415A", size = 16, face = "bold.italic"),
      legend.title = element_text(color = "#2E415A", size = 12, face = "bold"),
      legend.text = element_text(color = "#2E415A", size = 10),
      axis.title = element_text(color = "#2E415A", size = 14),
      axis.text = element_text(color = "#2E415A", size = 12))+
    scale_color_manual(values = c("#E0A151","#B04326"),
                       labels = c("Working on compassion", "Working on conscientiousness"))



```


### Changes in State-Level Conscientiousness

Also now realizing that it can be easier to filter datasets in the midst of making a plot instead of feeling like you have to do everything up front!

```{r plot-4}

ggplot(filter(VPCstate, !is.na(intervention)), aes(x = StateLinearTime, y = stateconscientiousness, group = id, color = intervention)) +
  geom_line() +
  labs(title = "Changes in state-level conscientiousness", subtitle = "by intervention group", x = "Week", y = "State Conscientiousness (1-5)", color = "Intervention group") +
   theme(plot.title = element_text(color = "#2E415A", size = 18, face = "bold"),
          plot.subtitle = element_text(color = "#2E415A", size = 16),
          strip.text.x = element_text(color = "#2E415A", size = 16, face = "bold.italic"),
      legend.title = element_text(color = "#2E415A", size = 12, face = "bold"),
      legend.text = element_text(color = "#2E415A", size = 10),
      axis.title = element_text(color = "#2E415A", size = 14),
      axis.text = element_text(color = "#2E415A", size = 12))+
    scale_color_manual(values = c("#E0A151","#B04326"),
                       labels = c("Working on compassion", "Working on conscientiousness"))



```

Oh my. This is great proof of the importance of having many different state assessments over the course of at least 5 weeks!! So much fluctuation!

### Changes in State-Level Compassion


```{r plot-5}

ggplot(filter(VPCstate, !is.na(intervention)), aes(x = StateLinearTime, y = statecompassion, group = id, color = intervention)) +
  geom_line() +
  labs(title = "Changes in state-level compassion", subtitle = "by intervention group", x = "Week", y = "State Compassion (1-5)", color = "Intervention group") +
   theme(plot.title = element_text(color = "#2E415A", size = 18, face = "bold"),
          plot.subtitle = element_text(color = "#2E415A", size = 16),
          strip.text.x = element_text(color = "#2E415A", size = 16, face = "bold.italic"),
      legend.title = element_text(color = "#2E415A", size = 12, face = "bold"),
      legend.text = element_text(color = "#2E415A", size = 10),
      axis.title = element_text(color = "#2E415A", size = 14),
      axis.text = element_text(color = "#2E415A", size = 12))+
    scale_color_manual(values = c("#E0A151","#B04326"),
                       labels = c("Working on compassion", "Working on conscientiousness"))



```

Again taking out the conscientiousness folks:

```{r plot-6}

ggplot(CompState, aes(x = StateLinearTime, y = statecompassion, group = id, color = intervention)) +
  geom_line() +
  labs(title = "Changes in state-level compassion", x = "Week", y = "State Compassion (1-5)", color = "Intervention group") +
   theme(plot.title = element_text(color = "#2E415A", size = 18, face = "bold"),
          plot.subtitle = element_text(color = "#2E415A", size = 16),
          strip.text.x = element_text(color = "#2E415A", size = 16, face = "bold.italic"),
      legend.title = element_text(color = "#2E415A", size = 12, face = "bold"),
      legend.text = element_text(color = "#2E415A", size = 10),
      axis.title = element_text(color = "#2E415A", size = 14),
      axis.text = element_text(color = "#2E415A", size = 12))+
    scale_color_manual(values = c("#E0A151","#B04326"),
                       labels = c("Working on compassion", "Working on conscientiousness"))



```

