---
title: "Portfolio 12: Webscraping Pitchfork"

---

For this portfolio, I decided to scrape Pitchfork (a music review website) to be able to work with data on what reviewers on the site have ranked the top 200 albums of the 2010s.



### Loading packages, checking permissions

```{r load-packages, message = FALSE}
library(tidyverse) 
library(skimr)
library(rvest)
library(robotstxt)
library(xml2)
library(dplyr)
library(tidyr)
```


```{r checking-permissions}
paths_allowed("https://pitchfork.com/features/lists-and-guides/")
```



## Top 200 Albums of the 2010s

### Prepping the dataset

I used SelectorGadget to figure out that I need to scrape ".heading-h3" for the ranking number and "h2" for a string with this format:

Artist: Album Name (Year)


```{r scrape-2010s}


library(rvest)
library(tidyverse)

url <- "https://pitchfork.com/features/lists-and-guides/the-200-best-albums-of-the-2010s/"
page <- read_html(url)

titlesartistsyears <- page %>% html_nodes("h2") %>% html_text()
rankings <- page %>% html_nodes(".heading-h3") %>% html_text()

albums_df <- tibble(titlesartistsyears = titlesartistsyears, rankings = rankings)


```

Now I'm going to work on cleaning up the data I've scraped.

First I need to separate the "titlesartistsyears" variable into three distinct variables.

When I first tried to do this, there were three cases that just wouldn't work with my code because they have extra parentheses in the title or artist name, so I decided to filter them out and then manually put them back in. 

```{r clean-strings-new}


# filter out rows 11, 34, and 106: these are problematic strings that are messing up my string separation because they have extra parentheses

albums_df_197 <- albums_df %>%
  filter(!row_number() %in% c(11, 34, 106))


# with filtered data frame: separate string into artist and rest of string
albums_df_197_separated <- albums_df_197 %>%
  separate(titlesartistsyears, c("artist", "rest"), sep = ": ") %>%
  # separate rest of string into album and year
  separate(rest, c("album", "year"), sep = " \\(|\\)")

# remove leading/trailing spaces
albums_df_197_separated <- albums_df_197_separated %>%
  mutate(across(everything(), str_trim))

# Now I need to add back in the 3 problematic strings.

new_row <- data.frame(artist = "Huerco S.", album = "For Those of You Who Have Never (and Also Those Who Have)", year = "2016", rankings = "190.")
albums_df_198 <- rbind(albums_df_197_separated, new_row)

new_row <- data.frame(artist = "(Sandy) Alex G", album = "DSU", year = "2014", rankings = "167.")
albums_df_199 <- rbind(albums_df_198, new_row)

new_row <- data.frame(artist = "Jai Paul", album = "Leak 04-13 (Bait Ones)", year = "2019", rankings = "95.")
albums_df_200 <- rbind(albums_df_199, new_row)

albums_df_200
```


Now I need to make sure all of my variables have the right formatting 

```{r fixing-variables}


# fixing the "ranking" variable

albums_df_200 <- albums_df_200 %>%
  mutate(ranking = as.numeric(sub("\\.$", "", rankings)))

albums_df_200 <- albums_df_200 %>%
select(-rankings)

# fixing the "year" variable so it is actually a year

albums_df_200 <- albums_df_200 %>%
  mutate(year = as.numeric(year))

albums_df_200
```

### Looking at the data

First, I think I want to create a new data frame that focuses on each discrete artist that has at least one album in the list. Specifically, I want to be able to look at average list ranking and number of albums in the list for each artist. 

```{r artist-centered-dataset}

albums_by_artist <- albums_df_200 %>%
  group_by(artist) %>%
  summarize(average_ranking = mean(ranking), 
            albums_in_list = n()) 


```

First I will sort this in ascending order by average ranking:

```{r lowest-average-ranking}
albums_by_artist_lowestavg <- albums_by_artist %>%
   arrange(average_ranking)

albums_by_artist_lowestavg
```


Wow I feel so vindicated!! Fiona Apple is one of my all-time favorite artists and she has the lowest average album ranking! (That is just from one album, but still. That album is the best album of the 2010s, IMHO.)

Now I will sort the dataset in descending order by number of albums in the list:

```{r most-albums-in-list}
albums_by_artist_listnumber <- albums_by_artist %>%
     arrange(desc(albums_in_list))

albums_by_artist_listnumber
```

Looks like Beyoncé and Kendrick Lamar are the only artists with 3 albums in the top 200. That checks. 

