---
title: "Portfolio 3: Thesis Part 3 (Observer-Level Analyses)"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```


This is part 3 of the analyses for my master's thesis project, which was titled "Effects of a Short-Term Volitional Intervention Promoting Moral and Non-Moral Personality Change." Here is the abstract of my thesis for some context:

> People can intentionally change specific personality traits. However, it is unclear whether such interventions are effective at changing moral character traits. The present study (total N = 179), which utilized two active intervention groups for mutual comparisons, examined whether a 5-week goal-setting intervention could lead to trait change for a moral trait, compassion (n = 87), or the Big 5 trait of conscientiousness (n = 92). Participants self-selected into one condition after getting feedback on where they stood on these traits in relation to the population norm and completing a brief reflection task. They set implementation intentions and reflected on their progress every 3 days for the duration of the intervention. State assessments were collected every 3 days in addition to self- and observer-report trait assessments at pre- and post-test, with follow-up self-report assessments 2 weeks post-program. Results from analyses of self-report trait and state assessments indicated that participants in both intervention groups were able to successfully change their personalities in the short term, providing preliminary evidence that implementation intention-based volitional personality change interventions can be useful tools for both moral and non-moral trait change. 


### Loading Packages and Data


```{r load-packages-data}


library(foreign)

VPCobserver = read.spss("observer data wide clean 3.23.23.sav",
                      to.data.frame = TRUE,
                      reencode = TRUE,
                      use.value.labels = FALSE)

VPCobserver_wider = read.spss("observer data even wider clean 3.23.23.sav",
                      to.data.frame = TRUE,
                      reencode = TRUE,
                      use.value.labels = FALSE)


library(nlme)
library(lme4)
library(lmerTest)
library(psych)
library(multilevel)
library(tidyverse)
library(broom)

VPCobserver$id <- as.factor(VPCobserver$id)
VPCobserver$intervention <- as.factor(VPCobserver$intervention)

ConsObserver <- filter(VPCobserver, intervention == 1)
CompObserver <- filter(VPCobserver, intervention == 0)

VPCobserver_wider$id <- as.factor(VPCobserver_wider$id)
VPCobserver_wider$intervention..00 <- as.factor(VPCobserver_wider$intervention..00)

ConsObserver_wider <- filter(VPCobserver_wider, intervention..00 == 1)
CompObserver_wider <- filter(VPCobserver_wider, intervention..00 == 0)



```

----------------


### Hypothesis 9

Hypothesis 9: I hypothesized that those in the **conscientiousness** group will have ***higher levels of trait conscientiousness*** at posttest compared to pretest as reported by close others acting as informants. 

Starting with an intercept only model:

```{r trait-conscienctiousness-null}
ConsObs <- lmer(avg_cons ~  1 +
                (1 | id), 
              data = ConsObserver,
              REML = FALSE,
              na.action = "na.omit")

summary(ConsObs)
confint(ConsObs)
rand(ConsObs)
AIC(ConsObs)
BIC(ConsObs)
```

Then fitting a linear model (I can't fit a logarithmic model with the observer data because there were only two observer data collection points):

```{r trait-conscienctiousness-linear}
# ConsObs <- lmer(avg_cons ~  LinearTime +
#                 (1 + LinearTime | id), 
#               data = ConsObserver,
#               REML = FALSE,
#               na.action = "na.omit")
# 
# summary(ConsObs)
# confint(ConsObs)
# rand(ConsObs)
# AIC(ConsObs)
# BIC(ConsObs)
```

Attempting to run this code gave me the following error code:

"Error: number of observations (=74) <= number of random effects (=88) for term (1 + LinearTime | id); the random-effects parameters and the residual variance (or scale parameter) are probably unidentifiable"

Here, I realized that I could not run mixed-effects linear models on the observer data because there are more random effects than there are participants (i.e., I didn't have a large enough informant sample size). I decided to run paired t-tests instead. 

```{r observer-t-test-cons-group}

cons_ttest <- t.test(ConsObserver_wider$avg_cons.5.00, ConsObserver_wider$avg_cons..00, paired = TRUE, alternative = "greater")
tidy(cons_ttest)

```


#### Conclusion:

> Hypothesis 9 was not fully supported: observers reported changes in trait conscientiousness in the expected direction for participants in the conscientiousness group, (t(30) = 1.48, d = 0.27), but these results were only marginally significant (p = .08). 


### Hypothesis 10 

Hypothesis 10: I hypothesized that those in the **compassion** group will have ***higher levels of trait compassion*** at posttest compared to pretest as reported by close others acting as informants. 

Starting with an intercept only model:

```{r trait-compassion-null}
CompObs <- lmer(avg_comp ~  1 +
                (1 | id), 
              data = CompObserver,
              REML = FALSE,
              na.action = "na.omit")

summary(CompObs)
confint(CompObs)
rand(CompObs)
AIC(CompObs)
BIC(CompObs)
```

Then linear:

```{r trait-avg_comp-linear}
# CompObs <- lmer(avg_comp ~  LinearTime +
#                 (1 + LinearTime | id), 
#               data = CompObserver,
#               REML = FALSE,
#               na.action = "na.omit")
# 
# summary(CompObs)
# confint(CompObs)
# rand(CompObs)
# AIC(CompObs)
# BIC(CompObs)
```
I received this error message:

"Error: number of observations (=59) <= number of random effects (=76) for term (1 + LinearTime | id); the random-effects parameters and the residual variance (or scale parameter) are probably unidentifiable"


Again, same problem as before. Below is the relevant paired-samples t-test:

```{r observer-t-test-comp-group}

comp_ttest <- t.test(CompObserver_wider$avg_comp.5.00, CompObserver_wider$avg_comp..00, paired = TRUE, alternative = "greater")
tidy(comp_ttest)

```

#### Conclusion:

> Hypothesis 10 was not supported: observers reported changes in trait compassion in the expected direction for participants in the compassion group, (t(20) = 0.73, d = 0.16), but these results were not significant (p = .24). 


